Modalities for Location Virtualization

### 2022-09-23 relwork.tex
some todos and some notes about Kolanski and Klein

Kolanski and Klein's points-to assertions do not model that page table entries for nearby addresses typically \emph{share} entries in higher layers of the page tables,\todo{actually, not sure, need to confirm} a key limitation that our definitions avoid.
They do not consider virtuallv address aliasing (another source of sharing physical resources), which we\todo{todo!} support.
They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
The mapping write is verified in an ad hoc way by unfolding the machine semantics, because the logic lacks proper reusable rules for
updating page tables.  
Critically, they do not consider
manipulating \emph{other} address spaces besides the one in use (essential for managing
processes).
So their means for sharing access to intermediate page table levels does not extend across address spaces, and there is no way to share data pages between address spaces.  But every major OS includes widely-used facilities for sharing pages between processes (e.g., UNIX/POSIX \texttt{MAP\_SHARED}), and until the recent Spectre~\cite{kocher2018spectre} and Meltdown~\cite{lipp2018meltdown} attacks, FreeBSD, Solaris, and others shared the kernel mapping across processes~\cite{mckusick2014design,mcdougall2006solaris,wininternals6a,wininternals6b} (they still do so for performance on unaffected hardware; e.g., Meltdown affects Intel but not AMD~\cite{amdmeltdown}).

>>>>

Kolanski and Klein's points-to assertions do not model that page table entries for nearby addresses typically \emph{share} entries in higher layers of the page tables,\todo{actually, not sure, need to confirm} a key limitation that our definitions avoid.
They do not consider virtuallv address aliasing (another source of sharing physical resources), which we\todo{todo!} support.
They verify only one address space manipulation: mapping a single unmapped page into the current address space (in both papers).
The mapping write is verified in an ad hoc way by unfolding the machine semantics, because the logic lacks proper reusable rules for
updating page tables.

They also define their virtual points-to assertions such that a virtual points-to $p\mapsto_\mathsf{v} a$ owns the full lookup path to virtual address $p$. This means that given two virtual points-to assertions at the same time, such as $p\mapsto_\mathsf{v}a \ast p'\mapsto_\mathsf{v}b$, the memory locations traversed to translate $p$ and $p'$ must be disjoint. This means the logic has a peculiar limit on how many virtual points-to assertions can coexist in a proof. Since page tables fan out, the bottleneck is the number of entries in the root table. For their 32-bit ARMv6 example, the top-level address is still 4Kb (4096 bytes), and each entry (consumed entirely by a virtual points-to in their scheme) is 4 bytes, so they have a maximum of 1024 virtual points-tos in their ARMv6 configuration. (They do formulate their logic over an abstract model, but every architecture would incur a similar limitation.) Na\"ively transferring their model to x86-64 4-level tables would yield a limit of 512 assertions (also a 4Kb root page, but 8-byte entries). Our initial prototype worked around this limitation using fractional permissions on the intermediate entries, so entries in the L1 table were split 512 ways, entries in the L2 were split $512^2$, and so on --- but this results in problems when considering identity mappings of page tables (which Kolanski and Klein's approach also does not support), where the data portion of a virtual points-to to a an intermediate table entry must also coexist with the shares for the ``traversal'' portion of the virtual points-tos those entries support. In principle knowing how many such identity-like mappings might exist could allow shifting the fractions, but this quickly becomes very complex.
Our approach\todo{v2} of separating the physical support for a virtual address mapping from the \emph{guarantee of} physical support for the virtual address avoids these issues.

Of critical relevance to the distinction between our work and theirs, they do not consider
manipulating \emph{other} address spaces besides the one in use (essential for managing
processes).
So their means for sharing access to intermediate page table levels does not extend across address spaces, and there is no way to share data pages between address spaces.  But every major OS includes widely-used facilities for sharing pages between processes (e.g., UNIX/POSIX \texttt{MAP\_SHARED}), and until the recent Spectre~\cite{kocher2018spectre} and Meltdown~\cite{lipp2018meltdown} attacks, FreeBSD, Solaris, and others shared the kernel mapping across processes~\cite{mckusick2014design,mcdougall2006solaris,wininternals6a,wininternals6b} (they still do so for performance on unaffected hardware; e.g., Meltdown affects Intel but not AMD~\cite{amdmeltdown}).
